\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2022-12-05 main.tex
John's comments

@@

\begin{abstract}
In recent years, many programming communities have been undergoing a transition to writing strongly typed programs. To benefit from type-safe codebases, programmers must first overcome the challenge of resolving type errors where text messages can often be unhelpful and misleading. Modern programming tools do not concentrate on making type errors easy to understand. We demonstrate \chameleon{}, a type debugging tool capable of displaying the full context of a type error and allowing programmers to search for where the type error actually occurs based on their prior knowledge and to incrementally verify their assumptions. \chameleon{} was developed through a user-centered, iterative process with four participatory experiments. Through the experiments, we showed that programmers using \chameleon{} fix type errors faster than traditional text-based error messages.  This difference is more significant in harder tasks. Further, programmers actively using \chameleon{} interactive features are shown to be more efficient in fixing type errors than simply reading the type error output.
\end{abstract}

@@

\bibliographystyle{IEEEtran}
\bibliography{icpc}

>>>>

% Comment out second line to disable.
\newcommand{\todo}[1]{}
\renewcommand{\todo}[1]{{\color{red} TODO: {#1}}}

@@

\begin{abstract}

\todo{I would start with saying untyped langs popular e.g. Python and the need to address problems this brings...}

In recent years, many programming communities have been undergoing a transition to writing strongly typed programs. To benefit from type-safe codebases, programmers must first overcome the challenge of resolving type errors where text messages can often be unhelpful and misleading. Modern programming tools do not concentrate on making type errors easy to understand. We demonstrate \chameleon{}, a type debugging tool capable of displaying the full context of a type error and allowing programmers to search for where the type error actually occurs based on their prior knowledge and to incrementally verify their assumptions. \chameleon{} was developed through a user-centered, iterative process with four participatory experiments. Through the experiments, we showed that programmers using \chameleon{} fix type errors faster than traditional text-based error messages.  This difference is more significant in harder tasks. Further, programmers actively using \chameleon{} interactive features are shown to be more efficient in fixing type errors than simply reading the type error output.
\end{abstract}

@@

\bibliographystyle{IEEEtran}
\bibliography{ICPC}

### 2022-12-05 sections/discussion.tex
John's comments

\section{Discussion}
This paper presents the interactive type debugging tool \chameleon{} and charts the evolution of its design across several iterations in response to user evaluation and feedback, as well as examining the effectiveness of the general approach compared to traditional static type error messages. We found that programmers using \chameleon{} type compare tool are able to debug errors faster than using traditional text-based error messages. This effect is shown more clearly when the task is not trivial. We found that programmers who actively use ChameleonIDE's interactive features (candidate expression cards and deduction steps) are more efficient in fixing type errors than simply reading the type error output. In this section, we will discuss a few interpretations of the results.

From the results of user studies 1 and 2, we observed that the choice of debugging tool had little effect on how fast programmers solve type errors. Conversely, when facing more realistic problems (longer source code, error locations more scattered), programmers are more effective using \chameleon{}. We believe one factor of this effectiveness of \chameleon{} is to reduce the amount of reading time. Jbara \cite{jbara2017programmers} showed that reading source code is generally the initial step of solving programming problems and is done in several passes. Although traditional compiler error message tools initially show fewer locations, it is unreliable, meaning that programmers have to expand the reading span without clear guidance. In contrast, Chameleon shows more error locations initially. However, the completeness of error locations assures programmers which part of the source code can be safely skipped.

@@

>>>>

\section{Discussion}

\todo{Suggest try and strucuture this a bit more e.g. each para bold the key lesson @ start of para. Can you group into implications for research/implications for practice to make more actionable??}

This paper presents the interactive type debugging tool \chameleon{} and charts the evolution of its design across several iterations in response to user evaluation and feedback, as well as examining the effectiveness of the general approach compared to traditional static type error messages. We found that programmers using \chameleon{} type compare tool are able to debug errors faster than using traditional text-based error messages. This effect is shown more clearly when the task is not trivial. We found that programmers who actively use ChameleonIDE's interactive features (candidate expression cards and deduction steps) are more efficient in fixing type errors than simply reading the type error output. In this section, we will discuss a few interpretations of the results.

From the results of user studies 1 and 2, we observed that the choice of debugging tool had little effect on how fast programmers solve type errors. Conversely, when facing more realistic problems (longer source code, error locations more scattered), programmers are more effective using \chameleon{}. We believe one factor of this effectiveness of \chameleon{} is to reduce the amount of reading time. Jbara \cite{jbara_how_2015} showed that reading source code is generally the initial step of solving programming problems and is done in several passes. Although traditional compiler error message tools initially show fewer locations, it is unreliable, meaning that programmers have to expand the reading span without clear guidance. In contrast, Chameleon shows more error locations initially. However, the completeness of error locations assures programmers which part of the source code can be safely skipped.

@@

\subsection{Threats to validity}

\todo{Put threats BEFORE Discussion and at end Eval section?}

\todo{These are too terse/unclear - need expanding. Suggest classify as internal / external threats etc .  What about background and number of participants as threats??}

One consequence of our recruiting approach is it is harder for previous participants to enter a later study. To minimize the lurking variable of previous experience, we use new code challenges in every study and conduct trial runs in every study to bring new participants up to speed.

Conducting studies remotely and unsupervised is that intervening when users encounter usability issues mid-study is impossible. To ensure there is no major usability issue that could lower the quality of data collection, we conducted cognitive walkthroughs and sandbox pilots before running each study.

### 2022-12-05 sections/evaluation.tex
John's comments

Three studies were conducted to answer new research questions or rephrase existing ones. These studies are grouped into 2 phases, and each tested a different version of \chameleon{} with human participants. The number of participants and their Haskell experience, as well as key research questions and conclusions, were shown in figure \ref{fig:timeline}.

@@

One consequence of our recruiting approach is it is harder for previous participants to enter a later study. To minimize the lurking variable of previous experience, we use new code challenges in every study and conduct trial runs in every study to bring new participants up to speed.

\subsubsection*{\textbf{Experiment setting}}
The experiments took place remotely and unsupervised. Participants took the study online via a web browser and at the physical venue of their choosing. All user studies use a web-based debugging environment developed by the authors. Conducting the studies online helped us avoid variation when performing tasks in unfamiliar places and using different setups. The downside is that intervening when users encounter usability issues mid-study is impossible. To ensure there is no major usability issue that could lower the quality of data collection, we conducted cognitive walkthroughs and sandbox pilots before running each study.

@@

After completing all the tasks, users are prompted to complete a debriefing survey. The survey questions include their Haskell experience and feedback on the tools and feature set participants used during the study.

We used a browser session recording tool~\cite{openreplay_openreplay_2022} to record the study sessions. This provides us with a "poor man's video recording"  to find usability issues in the study and recognize general patterns. The only reason this format is inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.

One other use of the session recording is for outlier trimming.   identify participants who left the computer for a long period of time. We first ran a statistical analysis to identify the long gap (2sd) between program logs for each participant. Then, we manually checked with the session recording to verify there was actually no user input (mouse movement, scrolling).

@@

\subsubsection{\textbf{\chameleon{} 1}}  
\chameleon{} 1 is an earlier iteration of \chameleon{} development. This version features the type inference engine that recovers most concrete types after type errors occur and a minimal set of debugging features. Key features in \chameleon{} 1 include showing two alternative types for the ill-typed expression,  providing all possible locations that contribute to the type error, and dividing possible locations into two groups based on which alternative type can be inferred. In terms of features, \chameleon{} 1 is equivalent to \chameleon{} 2 in basic mode.

@@

Two user studies (1a & 1b) were conducted to compare the effectiveness of solving type errors using \chameleon{} 1 and GHC compiler error messages. We choose GHC compiler error messages as the baseline because it is the canonical tool for working with type errors in Haskell. Although high-level tools like Haskell Language Server exist, they generally relay the GHC error messages verbatim. The procedure of these studies is shown in figure \ref{fig:procedure-1}.

Eight tasks were given in both studies. In study 1a, the tasks were taken from exercises from the Haskell programming units. These tasks cover a variety of type errors. However, these tasks, as shown from the result and feedback of user study 1, were too simple to fully evaluate the two options. In the second study, the tasks are sourced from the top 20 Haskell topics on GitHub~\cite{github_github_2022}. The authors manually added type errors.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/procedure-1.pdf}
    \caption{}
    \label{fig:procedure-1}
\end{figure}

@@

From the data collected during study 1a (\ref{fig:analysis-1a}), we were not able to identify which group performed better. For the trivial challenges we set for users, the individual differences are generally more significant than differences between treatments. The most common feedback (7 out of 35) the study 1a was that the tasks were too trivial to invite meaningful evaluation. One participant said, "Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."

One interesting observation is task 8, where the \chameleon{} group outperformed the GHC group more consistently and with a larger effect size (p-value = 0.002456, Wilcoxin signed test). The discrepancy is thought to be related to the nature of Task 8. It has a lengthier source file and involves more language features (abstract data types and high-level functions). GHC struggles to produce a relevant error message for this type of error.

In study 1b, the \chameleon{} group solved the type error faster (p =0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1. The authors are unclear about the cause of low performance from \chameleon{} in task 1. It is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity and ineffective training. For all the other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program.

In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}. "It's most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant. "I think \chameleon{}  does a much better job than GHC's error messages. I like that it shows the sources for the type judgments. This makes it quite easy to figure out how to rectify errors." reported another participant.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/procedure-2.pdf}
    \caption{}
    \label{fig:procedure-2}
\end{figure}

>>>>

Three studies were conducted to answer new research questions or rephrase existing ones. These studies are grouped into 2 phases, and each tested a different version of \chameleon{} with human participants. The number of participants and their Haskell experience, as well as key research questions and conclusions, were shown in figure \ref{fig:timeline}. \todo{Explain was a user study of the tool with practitioners?}

@@

\todo{Need more detail on participants - years experience, male vs female, etc}

\subsubsection*{\textbf{Experiment setting}}
The experiments took place remotely and unsupervised. Participants took the study online via a web browser and at the physical venue of their choosing. All user studies use a web-based debugging environment developed by the authors. Conducting the studies online helped us avoid variation when performing tasks in unfamiliar places and using different setups.

@@

After completing all the tasks, participants are prompted to complete a debriefing survey. The survey questions include their Haskell experience and feedback on the tools and feature set participants used during the study.

We used a browser session recording tool~\cite{openreplay_openreplay_2022} to record the study sessions. This provides us with a "poor man's video recording" to identify usability issues in the study and to recognize general patterns. The reason we found this format inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.

One use of the session recording is for outlier trimming.  We identify participants who left the computer for an extended period of time. We first ran a statistical analysis to identify the long gap (2 standard deviations) between actions using program logs for each participant. Then, we manually checked with the session recording to verify there was actually no user input (mouse movement, scrolling).

@@

\subsubsection{\textbf{\chameleon{} 1}}  
\chameleon{} 1 is an earlier iteration of \chameleon{} development. This version features the type inference engine that recovers most concrete types after type errors occur and a minimal set of debugging features. Key features in \chameleon{} 1 include showing two (or more) alternative types, showing all possible error locations, dividing possible error locations into groups based on alternative types, and concrete type restoration. In short, \chameleon{} 1 is equivalent to \chameleon{} 2 set to basic mode.

@@

Two  studies (1a \& 1b) were conducted to compare the effectiveness of solving type errors using \chameleon{} 1 and GHC compiler error messages. We choose GHC compiler error messages as the baseline because it is the canonical tool for working with type errors in Haskell. Although high-level tools like Haskell Language Server exist, they generally relay the GHC error messages verbatim. The procedure of these studies is shown in figure \ref{fig:procedure-1}.

Eight tasks were given in both studies. In study 1a, the tasks were taken from the exercises of the Haskell programming class in the authors' institute. These tasks cover a variety of type errors.  In the second study, the tasks are sourced from the top 20 Haskell topics on GitHub~\cite{github_github_2022}. The authors then manually added type errors into the program. In both studies, the type errors include simple mismatch, confusing syntax, missing instance, precedence and fixation, infinite types, and confusing list versus element. The categories of type errors to include in the tasks followed the common type errors in Tirronen's study \cite{tirronen_understanding_2015}.

The  studies (1a \& 1b) are designed to answer the research question:
\begin{itemize}
    \item \textbf{RQ1.} Do programmers solve type errors faster with \chameleon{} than GHC compiler error messages?
\end{itemize}

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/procedure-1.pdf}
    \caption{\todo{Does this figure really help at all?}}
    \label{fig:procedure-1}
\end{figure}

@@

From the data collected during study 1a (\ref{fig:analysis-1a}), we were not able to identify which group performed better. For the trivial challenges we set for users, the individual differences are generally more significant than differences between treatments. One interesting observation is task 8, where the \chameleon{} group outperformed the GHC group more consistently (p-value = 0.002456, Wilcoxin signed test). The discrepancy is thought to be related to the nature of Task 8. It has a lengthier source file and involves more language features (abstract data types and high-level functions). GHC struggles to produce a relevant error message for this type of error. We hypothesized that we may observe a more significant difference using tasks with lengthier and more realistic source code. This hypothesis is also supported by the most common feedback (7 out of 35) the study 1a was that the tasks were too trivial to invite meaningful evaluation. One participant said, "Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."

In study 1b, the \chameleon{} group solved the type error faster (p =0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1. For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity and ineffective training. For all the other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program. In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}. "It's most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant. "I think \chameleon{}  does a much better job than GHC's error messages. I like that it shows the sources for the type judgments. This makes it quite easy to figure out how to rectify errors." reported another participant.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/procedure-2.pdf}
    \caption{\todo{Does this figure really help at all?}}
    \label{fig:procedure-2}
\end{figure}

### 2022-12-05 sections/introduction.tex
John's comments

@@

Programmers find challenges in adopting typed languages in their practice. They tend to be verbose, hard to learn and generate bad error messages. 
One obstacle programmers face when using a statically typed language is understanding and resolving type errors~\cite{marceau_measuring_2011, tirronen_understanding_2015}, especially when dealing with modern type features such as polymorphic types and implicit typing. Studies show most statically typed languages produce unhelpful and even misleading type errors~\cite{}. Symptoms of bad error messages include cryptic language exposing internal constructs of the compiler and incomplete or wrong error locations. These usability issues pose challenges to learners and teachers of the languages. impede the wide adoption of statically typed languages, especially for programmers who are now accustomed to the mindset of writing un-typed programs.

This paper introduces \chameleon{}, an interactive debugging tool to visualize the relevant context of a type error (where it happens or could have happened, and which parts of code cause it) and allow programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve that ambiguity. Most noticeably, features (Section \ref{sub:type-compare}) provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature.  In addition, we integrated all the features into a unified debugging interface and provided functionality for programmers to enable and disable the features based on their preferences and debugging needs. \chameleon{} is open-source and is available in both web~\cite{shuai_fu_chameleon_2022} and desktop versions. The current implementation of \chameleon{} targets the Haskell 2010 language standard. However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques. Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages. Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak_history_2007}, and now can be found in most popular languages such as C\#~\cite{bill_wagner_constraints_2022}, Java~\cite{oracle_generic_2022}, and TypeScript~\cite{microsoft_documentation_2022}.

>>>>

\todo{As above, would start with motivation of dynamically typed langs e.g. JavaScript, Pythjon popularity but issues this brings...
}

Dynamically

@@

\todo{There are a lot of minor grammer fix-ups needed throughout...}
Programmers find challenges in adopting typed languages in their practice. \todo{I think this statement is plain wrong myself... 
 Need strong references to justify if going to say this to SE audience!:}They tend to be verbose, hard to learn and generate bad error messages. 
One obstacle programmers face when using a statically typed language is understanding and resolving type errors~\cite{marceau_measuring_2011, tirronen_understanding_2015}, especially when dealing with modern type features such as polymorphic types and implicit typing. Studies show most statically typed languages produce unhelpful and even misleading type errors~\cite{}. Symptoms of bad error messages include cryptic language exposing internal constructs of the compiler and incomplete or wrong error locations. These usability issues pose challenges to learners and teachers of the languages. impede the wide adoption of statically typed languages, especially for programmers who are now accustomed to the mindset of writing un-typed programs.

\todo{Way too long sentence ; confusing to reader - rephrase:}
This paper introduces \chameleon{}, an interactive debugging tool to visualize the relevant context of a type error (where it happens or could have happened, and which parts of code cause it) and allow programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve that ambiguity. Most noticeably, features provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature.  In addition, we integrated all the features into a unified debugging interface and provided functionality for programmers to enable and disable the features based on their preferences and debugging needs. \chameleon{} is open-source and is available in both web~\cite{shuai_fu_chameleon_2022} and desktop versions.  \todo{Introduce Haskell earlier?  Motivate WHY Haskell?} The current implementation of \chameleon{} targets the Haskell 2010 language standard. However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques. Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages. Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak_history_2007}, and now can be found in most popular languages such as C\#~\cite{bill_wagner_constraints_2022}, Java~\cite{oracle_generic_2022}, and TypeScript~\cite{microsoft_documentation_2022}.

### 2022-12-05 sections/related-work.tex
John's comments

\section{Related Work}
%%%  Todo: Discussing "Practical SMT-based Type Error Localization"
Researchers have long realized that incomprehensible type errors result as a consequence of standard approaches to the type inference process. Many studied [Citation here] the approach of finding all locations that contribute to a type error. Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}''  Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error. Internally labeled constraints and Minimal Unsatisfiable Set (MUS) manipulation are used to generate these slices. The language supported in Haack's work was a subset of Standard ML. The original Chameleon~\cite{stuckey_interactive_2003} used a similar technique but generate constraints in Constraint handling rules (CHR). As a result,  Chameleon supported advanced type-level features (type classes and functionally dependent types). The project also introduced the ability to query type information through a command line-style interface. Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies.

One direction to approach type error debugging is by suggesting solutions to fix the type error. Sheng Chen's works \cite{chen_counter-factual_2014, chen_efficient_2020, chen_improving_2022} on counterfactual typing are able to suggest changes to make the ill-typed program type-check. While this is a promising ground, the work is not evaluated with human participants. \cite{lerner_searching_2007} studied an approach of repeatedly modifying the AST (abstract syntax tree) and querying the compiler/type checker to know whether the new program is well-typed. This approach allows the algorithm to suggest syntax changes that could fix the type error. Benefiting from not needing a type system implementation and being language agnostic, the underlying idea was implemented and evaluated in two different languages with ten university students. While suggesting syntax changes is very intuitive in practice, this approach has limitations for teaching and learning purposes, as it is not able to explain why the type errors occur and why fixes are suggested. The lack of explanation is due to the inquiry into the underlying type-check being a binary state. Deep type-level analysis (typing rules and language features) was not communicated to the inference algorithm.

Another related approach is through ECEM (Enhanced compiler error message). Through a series of mixed-method studies, Prather showed \cite{prather_novices_2017} that ECEM has a positive result in understanding compiler errors. Decafe \cite{becker_effective_2016} is a tool that can rephrase the Java compiler error messages into their enhanced version. In a study of over 200 CS1 students, Decaf was shown to reduce overall errors in their coding practices. Berik proposed a framework \cite{barik_how_2018} for constructing compiler error messages based on argumentation theory. In this proposal,  Berik argued that error messages following a simple argumentation layout or an extended argumentation layout are more human-friendly.  These works show the significance of improving the language in the compiler error messages. Most principles and suggestions are followed in \chameleon{} in constructing error statements. However, they were not targeting type errors alone but compiler errors (some even include run time errors) in general. The nuances of type errors, such as alternative typing, were not considered.

% \begin{figure}[h]
%     \centering
%     \includegraphics[width=\linewidth]{images/original-chameleon.pdf}
%     \caption{
% Originally, Chameleon was a command-line tool that can list all the potential causes of a type error and display the two branches in two forms of highlight (right), achieved with ASCII colors. In addition, it allows a few commands (debug and explain) in the debugging shell to help debug type errors (left).}
%     \label{fig:original-chameleon}
% \end{figure}

% The ChameleonIDE error reporting The error message of ChameleonIDE closely
% follows the argumentation structure outlined by Titus. Basic mode represent a
% simple structure argument, where the claim "The expression e can have two
% conflicting types" and the possible type 1 and two serves as grounds. When
% inspecting each possible types, the type judgements are treated as claim, and it
% is supported by the grounds: "Inferred from the orange/blue highlights on the
% left side". On the other hand, the similar type error message in GHC "Couldn't
% match expect type T1 with actual type T2" is "a ground masquerading as a clam",
% Barik commented. The balanced mode and advanced mode serve as a way to elaborate
% arguments from simple arguments into extended arguments.

% Another related topic is type-explanation~\cite{yang1999explaining, jun2002explaining}. Yang~\cite{jun2002explaining} showed an alternative type inference system capable of producing a human-like text explanation for why expressions are assigned certain types. A good explanation is drawn from surveying how human experts explain types. The resulting algorithm $\mathcal{H}$, generates a succinct explanation of the type inference steps to avoid using internal constructs (such as type variables). The explanation system has the advantage of acting like a human expert. However, when presented as text-based output, explanation systems have the potential to become verbose when types are complex or variable names are long. In \chameleon{}, we attempt to address this problem by showing one step of explanation at a time and referring to variables instead of spelling out the full name.

Debugging using a GUI Interactive Development Environment has been the standard practice for a long time. Compared to a command line-based interface, a graphic user interface provides programming tools that can show information hierarchy, provide immediate feedback to changes, and display complex visualizations. Hazel Tutor \cite{potter_hazel_2020} is an interactive type-driven environment for the OCaml language. It can automatically fill type holes by suggesting template expressions (strategies as called by the authors) through a popup window. It also provides a cursor-based type inspector that allows programmers to query the types of different parts of the program. Whyline~\cite{ko_finding_2009} is a Java debugging system that allows a user to ask questions like "why does variable X have value Y." It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error.  These debugging tools are important motivations for developing \chameleon{}. However, they focus on different contexts of the debugging process. Java Whyline mainly tackles the problem of unintended run-time behavior, while Hazel Tutor specializes in interactive actions supporting type holes.

>>>>

\section{Related Work}

\todo{I would do a short Motivating Example section straight after intro showing example of untyped lang, problems it brings, and outline requirments for the proposed IDE to address it. Make sense??}

\todo{Do yoou need Related Work here - or put @ end of paper just before Summary - I tend to do the later now UNLESS its REALLY needed by reader - IMO could put @ end and get to your work/contributions much quicker}

\subsection{Finding all type error locations}
Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' Many studied \cite{stuckey_interactive_2003, haack_type_2004, pavlinovic_practical_2015, schilling_constraint-free_2012} the approach of finding all locations that contribute to a type error. Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error. Internally labeled constraints and Minimal Unsatisfiable Set (MUS) manipulation are used to generate these slices. The language supported in Haack's work was a subset of Standard ML. The original Chameleon~\cite{stuckey_interactive_2003} used a similar technique but generate constraints in Constraint handling rules (CHR). As a result,  Chameleon supported advanced type-level features (type classes and functionally dependent types). The project also introduced the ability to query type information through a command line-style interface. Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies. While finding all error locations is useful in comprehending type errors, it is only 1 of the 7 properties listed in the proposed manifesto of good type error reporting~\cite{yang_improved_2000}.

\subsection{Producing high-quality error explanation}
One weakness of compiler error messages, in general, is that they fail to explain the error in human language. As put in~\cite{barik_developers_2017}, "Error messages appear to take the form of natural language, yet are as difficult to read as source code."  A well-studied approach to producing better error explanations is through ECEM (Enhanced compiler error message). Through a series of mixed-method studies, Prather showed~\cite{prather_novices_2017} that ECEM has a positive result in understanding compiler errors. Decafe~\cite{becker_effective_2016} is a tool that can rephrase the Java compiler error messages into their enhanced version. In a study of over 200 CS1 students, Decaf was shown to reduce overall errors in their coding practices. Berik proposed a framework~\cite{barik_how_2018} for constructing compiler error messages based on argumentation theory. In this proposal,  Berik argued that error messages following a simple argumentation layout or an extended argumentation layout are more human-friendly.  These works show the significance of improving the language in the compiler error messages. Most principles and suggestions are followed in \chameleon{} in constructing error statements. However, they were not targeting type errors alone but compiler errors (some even include run time errors) in general. The nuances of type errors, such as alternative typing, were not considered. Moreover, explanation systems are designed specifically for novice users.

\subsection{Suggesting resolution}
One direction to approach type error debugging is by suggesting solutions to fix the type error. Sheng Chen's works~\cite{chen_counter-factual_2014, chen_efficient_2020, chen_improving_2022} on counterfactual typing are able to suggest changes to make the ill-typed program type-check. While this is a promising ground, the work is not evaluated with human participants. Lerner  studied an approach~\cite{lerner_searching_2007} of repeatedly modifying the AST (abstract syntax tree) and querying the compiler/type checker to know whether the new program is well-typed. This approach allows the algorithm to suggest syntax changes that could fix the type error. Benefiting from not needing a type system implementation and being language agnostic, the underlying idea was implemented and evaluated in two different languages with ten university students. While suggesting syntax changes is very intuitive in practice, this approach has limitations for teaching and learning purposes, as it is not able to explain why the type errors occur and why fixes are suggested. The lack of explanation is due to the inquiry into the underlying type-check being a binary state. Deep type-level analysis (typing rules and language features) was not communicated to the inference algorithm.

\subsection{Interactive Debugging}
Interactive debugging tools have been studied using the command line interface. Chameleon provides a certain level of interactive debugging~\cite{stuckey_interactive_2003}. The interaction is used to query different parts of the program. A recent study \cite{tsushima_type_2021} provided an interactive type debugging system that restores type information based on the output from CFT-based type inference.  Both tools allow programmers to change their assumptions about which expression is ill-typed. While both tools show a strong influence on theoretical development, they were not evaluated in realistic studies with human participants. In addition, command line-based debugging tools suffer the same limitation as other line-based tools, namely, 1 2 and 3.

A graphic user interface (GUI) provides programming tools with immediate feedback on changes and the ability to show complex visualizations. Many tools dedicate to improving the debugging experience. We list two. Hazel Tutor \cite{potter_hazel_2020} is an interactive type-driven environment for the OCaml language. It can automatically fill type holes by suggesting template expressions (strategies as called by the authors) through a popup window. It also provides a cursor-based type inspector that allows programmers to query the types of different parts of the program. Whyline~\cite{ko_finding_2009} is a Java debugging system that allows a user to ask questions like "why does variable X have value Y." It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error. These debugging tools are important motivations for developing \chameleon{}. However, they focus on different aspects of the debugging process. Java Whyline mainly tackles the problem of unintended run-time behavior, while Hazel Tutor specializes in development assistance supported by type holes.

% A few studies \cite{yang1999explaining, jun2002explaining} focus on explain type error in particular. Yang~\cite{jun2002explaining} showed an alternative type inference system capable of producing a human-like text explanation for why expressions are assigned certain types. A good explanation is drawn from surveying how human experts explain types. The resulting algorithm $\mathcal{H}$, generates a succinct explanation of the type inference steps to avoid using internal constructs (such as type variables). The explanation system has the advantage of acting like a human expert. However, when presented as text-based output, explanation systems have the potential to become verbose when types are complex or variable names are long.
